---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# GCalignR

![Build Status](https://travis-ci.org/mastoffel/GCalignR.svg?branch=master)

`GCalignR` provides simple functions to align gas-chromatography data based on retention times and plots to
evaluate the quality of the alignment.

You can install:

* the latest development version from github with

    ```R
    if (packageVersion("devtools") < 1.6) {
      install.packages("devtools")
    }
    devtools::install_github("mastoffel/GCalignR", build_vignettes = TRUE)
    ```
   

If you encounter bug or if you have any suggestions for improvement, just contact
martin.adam.stoffel[at]gmail.com or meinolf.ottensmann[at]web.de

## Get started with GCalignR

To get started read the vignette:  

```{r, eval = FALSE}
vignette("GCalignR_step_by_step", package = "GCalignR")
```
