---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# GCalignR

![Build Status](https://travis-ci.org/mastoffel/GCalignR.svg?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/GCalignR)](https://cran.r-project.org/package=GCalignR)
[![](http://cranlogs.r-pkg.org/badges/grand-total/GCalignR)](https://cran.r-project.org/package=GCalignR)


`GCalignR` provides simple functions to align peak lists obtained from Gas Chromatography Flame Ionization Detectors (GC-FID) based on retention times and plots to evaluate the quality of the alignment. The package supports any other one-dimensional chromatograpy technique that enables the user to create a peak list with at least one column specifing retention times.

Installing GCalignR:

* Get the latest development version from github with

```R
    if (!("devtools" %in% rownames(installed.packages()))) { 
    install.packages("devtools")
    } else if (packageVersion("devtools") < 1.6) {
    install.packages("devtools")
    }
    devtools::install_github("mastoffel/GCalignR", build_vignettes = TRUE)
```
* If the installation fails try installing the packages
devtools, ggplot2 and vegan with the commmands below (tested with a new installation of R 3.2.5)

```R
    if (!("devtools" %in% rownames(installed.packages()))) {
    install.packages("devtools")
    } else if (packageVersion("devtools") < 1.6) {
    install.packages("devtools")
    }
    install.packages("ggplot2",dependencies = TRUE)
    install.packages("vegan",dependencies = TRUE)
    devtools::install_github("mastoffel/GCalignR", build_vignettes = TRUE)    
```
### Get started with GCalignR

To get started read the vignette:  

```{r, eval = FALSE}
    vignette("GCalignR_step_by_step", package = "GCalignR")
```

If you encounter bugs or if you have any suggestions for improvement, just contact
meinolf.ottensmann[at]web.de

### Reference
[Ottensmann, M., Stoffel, M.A., Hoffman, J.I. GCalignR: An R package for aligning Gas-Chromatography data. *In review*.](https://doi.org/10.1101/110494)
