---
title: "GCalignR step by step"
author: "Martin A. Stoffel & Meinolf Ottensmann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
    %\VignetteIndexEntry{GCalignR step by step}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", cache = FALSE,
    fig.width = 5, fig.height = 4) # warning = FALSE
```


## GCalignR

GCalignR provides a simple means of aligning chromatograms from gas-chromatography (GC)
and gas-chromatography mass-spectrometry (GC-MS). Chromatograms of multiple samples are not
directly comparable as the retention times vary across samples due to sublte, random and often
unavoidable variation the GC-MS machine parameters [@pierce2005classification]. For studies
interested in broad patterns within chemical samples it is crucial that the same substances
will have the same retention times to compare them across samples. Suprisingly, within the
relatively young field of mammalian olfactory communication, most researchers seem to align
chromatograms manually (some citations) rather than using automated alignment algorithms, 
which are used by chemists or chemical plant biologists for a while (i am not sure about this,
needs citations). This might be due to (to what might it be due? No idea, but we also didn´t use
them obviously).

## The alignment algorithm

## Installation

The development version can be downloaded from [GitHub](https://github.com/mastoffel/GCalignR) with the following code:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("mastoffel/inbreedR", build_vignettes = TRUE)
```

The package provides a comprehensive documentation, check out the `GCalignR` help file.

```{r}
library("GCalignR")
```

```{r, eval = FALSE}
?GCalignR
```

## GCalignR contains the following functions: 

* `align_chromatograms`: Aligns gas-chromatography data by retention times. 
* `gc_heatmap`: Plots the results (and intermediate steps) of the alignment. With a heatmap,
the user can visually inspect the quality of the alignment and if necessary adjust the parameters.

## Step by step guide to aligning chromatograms with GCalignR

GC and GC-MS machines are coupled to specific software to look at the chromatograms and mass-spectra (i.e. Xcalibur).
These programs usually allow to extract the peaks of the chromatograms along with the areas under the peaks
and additional variables such as the peak height. The aim of GCalignR is to align these extracted peaks across individuals
using just their retention times (and not the mass-spectra). The simple underlying assumption is
that peaks with similar retention times represent the same substances. However, it is highly recommended
to verify this assumption by comparing also the mass-spectra (if available) of the substances of interest.

The alignment algorithm follows the following steps:

1. Given a reference chromatogram within the sample, all chromatograms are aligned to this reference
by means of shifting them to maximise shared peaks with the reference. This step corrects overall
differences in the retention times of chromatograms.

2. This step corrects unreliability of chromatograms on an individual peak basis and essentially
tries to minimise variation  within a retention time row. 

3. In a third step, retention time rows are merged if they have similar retention time means 
and non of the samples shows peaks in both or the rows (with the assumption that these two
rows represent a single substance)


## input file


## Literature


