---
title: "GCalignR step by step"
author: "Martin A. Stoffel & Meinolf Ottensmann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
    %\VignetteIndexEntry{inbreedR step by step}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", cache = FALSE,
    fig.width = 5, fig.height = 4) # warning = FALSE
```


## GCalignR

GCalignR provides a simple means of aligning chromatograms from gas-chromatography (GC)
and gas-chromatography mass-spectrometry (GC-MS). Chromatograms of multiple samples are not
directly comparable as the retention times vary across samples due to sublte, random and often
unavoidable variation the GC-MS machine parameters [@pierce2005classification]. For studies
interested in broad patterns within chemical samples it is crucial that the same substances
will have the same retention times to compare them across samples. Suprisingly, within the
relatively young field of mammalian olfactory communication, most researchers seem to align
chromatograms manually (some citations) rather than using automated alignment algorithms, 
which are used by chemists or chemical plant biologists for a while (i am not sure about this,
needs citations). This might be due to (to what might it be due? No idea, but we also didn´t use
them obviously).

## The alignment algorithm

## Installation

The development version can be downloaded from [GitHub](https://github.com/mastoffel/GCalignR) with the following code:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("mastoffel/inbreedR", build_vignettes = TRUE)
```

The package provides documentation for every function. To get an overview, just look at `GCalignR`'s help file.

```{r}
library("GCalignR")
```

```{r, eval = FALSE}
?GCalignR
```

## GCalignR contains the following functions: 

* `align_chromatograms`: Aligns gas-chromatography data by retention times. 
* `gc_heatmap`: Plots the results (and intermediate steps) of the alignment. With a heatmap,
the user can visually inspect the quality of the alignment and if necessary adjust the parameters.

## Step by step guide to aligning chromatograms with GCalignR

# input file
The 

## Literature


